---
title: "hw 3"
author: "cathy"
date: "2024-10-13"
output: github_document
---

Load the package that needed for this hw.

```{r, echo = FALSE}
library(tidyverse)
library(ggplot2)
```






## Problem 2

import and clean "nhanes_covar" data

```{r}
 covar_df =  
   read_csv("data/nhanes_covar.csv", 
             na = c("NA",".",""),
             skip = 4
            ) |> 
   janitor::clean_names()
```

import and clean "nhanes_accel" data
create a total activity variable for each participant

```{r}
 accel_df =  
   read_csv("data/nhanes_accel.csv", 
             na = c("NA",".","")) |> 
   janitor::clean_names() |>
   mutate(total_activity = 
            rowSums(across(starts_with("min")), na.rm = TRUE)) |> 
   relocate(total_activity)
```


Merge the 2 dataset into 1
```{r}
 merged_df =
   covar_df|> 
   left_join(accel_df, by = "seqn") |> 
   mutate(
    sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
    education = factor(education, levels = c(1, 2, 3), labels = c("Less than High School", "High School Equivalent", "More than High School"))
  ) |> 
   filter(age >= 21, !is.na(sex), !is.na(bmi), !is.na(education)) |> 
   relocate(seqn, sex, age, bmi, education, total_activity)

```


Create a visualization of the age distributions for men and women in each education category

```{r}
edu_bysex_table =
  merged_df |> 
  group_by(sex, education) |> 
  summarise(count = n(), .groups = 'drop')
```

create histogram and box plot
```{r}
edu_bysex_plot =
  ggplot(merged_df, aes(x = age, fill = sex)) +
  geom_histogram(position = "dodge", bins = 10) +
  facet_grid(sex ~ education) +
  labs(title = "Age Distribution by Sex and Education",
       x = "Age",
       y = "Count") +
  theme_minimal()

print(edu_bysex_plot)
```

```{r}
age_boxplot = ggplot(merged_df, aes(x = education, y = age, fill = sex)) +
  geom_boxplot() +
  facet_wrap(~ sex) +
  labs(title = "Age Distribution by Sex and Education",
       x = "Education",
       y = "Age") +
  theme_minimal()

print(age_boxplot)
```

Comment:



Plot these total activities (y-axis) against age (x-axis)
compare men to women and have separate panels for each education level

```{r}
 activity_plot =
   ggplot(merged_df, aes(x = age, y = total_activity, color = sex)) +
   geom_point(alpha = 0.5) +  
   geom_smooth(se = FALSE) + 
   facet_wrap(~ education, scales = "free_y") + 
   labs(title = "Total Daily Activity by Age, Sex, and Education Level",
        x = "Age",
        y = "Total Activity",
        color = "Sex") +
  theme_minimal() + 
  theme(legend.position = "bottom")

print(activity_plot)
```

Comment:


Make a three-panel plot that shows the 24-hour activity time 
 courses for each education level and use color to indicate sex

```{r}
long_accel_df =
  merged_df |> 
  pivot_longer(
    cols = starts_with("min"),
    names_to = "minute",
    names_prefix = "min",
    values_to = "activity_level"
  ) |> 
  mutate(
    minute = as.numeric(sub("min", "", minute)) / 1440 * 24 
  )
```


```{r}
 time_course_plot =
   ggplot(long_accel_df, aes(x = minute, y = activity_level, color = sex)) +
   geom_line(alpha = 0.3, size = 0.5) + 
   geom_smooth(se = FALSE, span = 0.2) + 
   facet_wrap(~ education, scales = "free_y") + 
   labs(
        title = "24-Hour Activity Time Course by Education Level and Sex",
        x = "Time of Day (hours)",
        y = "Activity Level",
        color = "Sex") +
   theme_minimal() +  
   theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5))

print(time_course_plot)
```

Describe:












